cardinalR: Generating interesting high-dimensional data structures

### 2025-10-13 paper-cardinalR.Rmd
Added benchmarking clustering algorithm

```{r load-libraries}
library(cardinalR)
library(tidyverse)
library(kableExtra)
library(geozoo)
library(patchwork)
library(colorspace)
library(crosstalk)
```

@@

```{r triginometric-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `crescent` (a1-a3), `curvycylinder` (b1-b3), `sphericalspiral` (c1-c3), `helicalspiral` (d1-d3), `conicspiral` (e1-e3), and `nonlinear` (f1-f3) data.", fig.width=12, fig.height=16, out.width="80%", fig.align='center', fig.pos="H"}

@@

positions <- geozoo::simplex(p=4)$points
positions <- positions * 0.3 #positions * 0.8

@@

nldr1 <- tsne_data |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("a", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.text = element_text(size = 10))

@@

nldr2 <- umap_data |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("b", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.text = element_text(size = 10))

@@

nldr3 <- phate_data |>
  ggplot(aes(x = PHATE1,
             y = PHATE2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("c", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.text = element_text(size = 10))

@@

nldr4 <- trimap_data |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("d", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.text = element_text(size = 10))

@@

nldr5 <- pacmap_data |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("e", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.text = element_text(size = 10))

@@

nldr6 <- pca_data |>
  ggplot(aes(x = pca1,
             y = pca2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("f", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.text = element_text(size = 10)) 
```

@@

nldr1 + nldr2 + nldr3 +
  nldr4 + nldr5 + nldr6 +
  plot_layout(ncol = 3, guides = "collect")
```

@@

```{r}
library(fpc)

# Extract high-dimensional data
data <- five_clusts[, -5]
true_labels <- as.numeric(gsub("cluster", "", five_clusts$cluster))

@@

```{r}
library(mclust)

# Model-based clustering
mclust_res <- Mclust(data, G = 5, modelNames = "VVV")$classification
```

@@

# Summarize key performance metrics
data.frame(
  Method = c("K-means", "Hierarchical", "Model-based"),
  Avg_Silhouette = c(comp_kmeans$avg.silwidth, comp_hclust$avg.silwidth, comp_mclust$avg.silwidth),
  Dunn_Index = c(comp_kmeans$dunn, comp_hclust$dunn, comp_mclust$dunn),
  Entropy = c(comp_kmeans$entropy, comp_hclust$entropy, comp_mclust$entropy),
  Corrected_Rand = c(comp_kmeans$corrected.rand, comp_hclust$corrected.rand, comp_mclust$corrected.rand)
)

>>>>

```{r load-libraries}
library(cardinalR)
library(tidyverse)
library(kableExtra)
library(geozoo)
library(patchwork)
library(colorspace)
library(crosstalk)
library(fpc)
library(mclust)
```

@@

```{r triginometric-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `crescent` (a1-a3), `curvycylinder` (b1-b3), `sphericalspiral` (c1-c3), `helicalspiral` (d1-d3), `conicspiral` (e1-e3), and `nonlinear` (f1-f3) data.", fig.width=12, fig.height=16, out.width="100%", fig.align='center', fig.pos="H"}

@@

positions <- geozoo::simplex(p=4)$points
positions <- positions * 0.3

@@

nldr1 <- tsne_data |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("a", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

nldr2 <- umap_data |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("b", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

nldr3 <- phate_data |>
  ggplot(aes(x = PHATE1,
             y = PHATE2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("c", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

nldr4 <- trimap_data |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("d", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

nldr5 <- pacmap_data |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("e", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.position = "none")

@@

nldr6 <- pca_data |>
  ggplot(aes(x = pca1,
             y = pca2,
             color = cluster))+
  geom_point(alpha=0.1, size=1) +
  interior_annotation("f", c(0.08, 0.93)) + 
  scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(aspect.ratio = 1,
        legend.position = "none") 
```

@@

nldr1 + nldr2 + nldr3 +
  nldr4 + nldr5 + nldr6 +
  plot_layout(ncol = 3)
```

@@

To further evaluate the structure of the generated data, we benchmarked three clustering algorithms: **$k$-means** [Chapter 20 of @boehmke2019], **hierarchical** [@murtagh2012], and **model-based clustering** [@chris2002; @scrucca2023] using the simulated dataset. The model-based clustering was performed with the `"VVV"` covariance structure, allowing each cluster to vary in volume, shape, and orientation. Cluster validity statistics were computed using the `cluster.stats()` function from the `fpc` package [@christian2024].

```{r}
# Extract high-dimensional data
data <- five_clusts[, -5]
true_labels <- as.numeric(gsub("cluster", "", five_clusts$cluster))

@@

```{r}
# Model-based clustering
mclust_res <- Mclust(data, G = 5, modelNames = "VVV")$classification
```

@@

# Summarize key performance metrics
summary_clust <- tibble::tibble(
  Metric = c("k-means", "Hierarchical", "Model-based"),
  `wb.ratio` = round(c(comp_kmeans$wb.ratio, comp_hclust$wb.ratio, comp_mclust$wb.ratio), 2),
  `Dunn Index` = round(c(comp_kmeans$dunn, comp_hclust$dunn, comp_mclust$dunn), 2),
  `Corrected rand` = round(c(comp_kmeans$corrected.rand, comp_hclust$corrected.rand, comp_mclust$corrected.rand), 2),
  VI = round(c(comp_kmeans$vi, comp_hclust$vi, comp_mclust$vi), 2)
)
```

```{r summaryclust-tb-html, eval=knitr::is_html_output()}
summary_clust |> 
  kable(caption = "cardinalR branching data generation functions", col.names = c("Metric", "wb.ratio", "Dunn_Index", "Corrected_Rand", "VI"), table.pos = "H") 
```

```{r summaryclust-tb-pdf, eval=knitr::is_latex_output()}
summary_clust |> 
  kable(caption = "cardinalR branching data generation functions", format="latex", col.names = c("Metric", "wb.ratio", "Dunn_Index", "Corrected_Rand", "VI"), booktabs = T, table.pos = "H")  |>
  column_spec(1, width = "2.5cm") |>
  column_spec(2, width = "2cm") |>
  column_spec(3, width = "2cm") |>
  column_spec(4, width = "2cm") |>
  column_spec(5, width = "2cm")
```

Overall, all methods produced similar compactness and separation, as reflected by the *within–between cluster ratios (wb.ratio)* and *Dunn indices*. However, the **model-based clustering** achieved the highest *Corrected Rand Index* ($0.75$) and lowest *Variation of Information (VI)* ($0.65$), indicating the best recovery of the true underlying groups. In comparison, $k$-means and hierarchical clustering showed moderate agreement with the true labels. These findings demonstrate that mixture-based approaches can more effectively capture the heterogeneity of clusters in high-dimensional, non-spherical data.

### 2025-10-13 paper-cardinalR.tex
Added benchmarking clustering algorithm

\begin{verbatim}
positions <- geozoo::simplex(p=4)$points
positions <- positions * 0.3 #positions * 0.8

@@

\begin{verbatim}
#>         Method Avg_Silhouette  Dunn_Index  Entropy Corrected_Rand
#> 1      K-means       0.227417 0.008968602 1.551532      0.4223504
#> 2 Hierarchical       0.227417 0.008968602 1.551532      0.5029130
#> 3  Model-based       0.227417 0.008968602 1.551532      0.7525318
\end{verbatim}

>>>>

\begin{verbatim}
positions <- geozoo::simplex(p=4)$points
positions <- positions * 0.3

@@

To further evaluate the structure of the generated data, we benchmarked three clustering algorithms: \textbf{\(k\)-means} \citep[Chapter 20 of][]{boehmke2019}, \textbf{hierarchical} \citep{murtagh2012}, and \textbf{model-based clustering} \citep{chris2002, scrucca2023} using the simulated dataset. The model-based clustering was performed with the \texttt{"VVV"} covariance structure, allowing each cluster to vary in volume, shape, and orientation. Cluster validity statistics were computed using the \texttt{cluster.stats()} function from the \texttt{fpc} package \citep{christian2024}.

\caption{\label{tab:summaryclust-tb-pdf}cardinalR branching data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}>{\raggedleft\arraybackslash}p{2cm}}
\toprule
Metric & wb.ratio & Dunn\_Index & Corrected\_Rand & VI\\
\midrule
k-means & 0.61 & 0.01 & 0.42 & 1.32\\
Hierarchical & 0.61 & 0.01 & 0.50 & 1.15\\
Model-based & 0.61 & 0.01 & 0.75 & 0.65\\
\bottomrule
\end{tabular}
\end{table}

Overall, all methods produced similar compactness and separation, as reflected by the \emph{within--between cluster ratios (wb.ratio)} and \emph{Dunn indices}. However, the \textbf{model-based clustering} achieved the highest \emph{Corrected Rand Index} (\(0.75\)) and lowest \emph{Variation of Information (VI)} (\(0.65\)), indicating the best recovery of the true underlying groups. In comparison, \(k\)-means and hierarchical clustering showed moderate agreement with the true labels. These findings demonstrate that mixture-based approaches can more effectively capture the heterogeneity of clusters in high-dimensional, non-spherical data.
